Generalized Bradley-Terry models

### 2025-03-07 inst/jss_hyper3_real.Rnw
more work

About 22\% of matches in Table~\ref{oldfirm} end in a draw and {\tt
  hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this has a
number of desirable features.  Firstly, if we consider two players of
identical strengths and nullify the home ground effect (that is,
$\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
odds of a draw is simply $D$, regardless of team strength.  Secondly,
considering only won (that is, not score-drawn) games, the probability
of a home win is $\frac{\lambda}{1+\lambda}$, that is, the odds of a
home win is simply $\lambda$, independently of the player
strength. Such arguments give direct operational significance to
$\lambda$ and $D$.

\begin{table}
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + \lambda p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  multiplicative Bradley-Terry model\label{windrawloss} and a lambda thing}
\end{table}

@@

\begin{figure}[htbp]
  \begin{center}
<<makecontour, cache=TRUE>>=
n <- 20
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
contour(lambda, D, matrix(LL, n, n),levels = -2*(0:9), xlab=expression(lambda),ylab=expression(D))
points(maxv[1], maxv[2], pch=16, col='red')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate}
  \end{center}
\end{figure}

@@

The red bus-blue bus problem was originally stated by
\cite{chipman1960} and discussed by \cite{davidson1970} and
\cite[p153]{alvo2014}.  
by \cite{tverskey1972}.
The canonical problem is as follows. A traveller has a choice of going
to work by different transport methods.  Here, we can assume that the
traveller has strong but unknown views on different forms of transport
which might incorporate a random component.  If considering whether to
use a car or a bus, the colour of the bus is of no importance to him.
Such preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, say, $p_{C} = p_{RB}$ and also
indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\Prob(C| \left\lbrace C,RB\right\rbrace)=\frac{1}{3}$.  Various
solutions to this problem have been proposed.  \cite{tverskey1972},
for example, presents an approach involving overlapping aspects;

If we ask him to
rank his options, it is highly probable that he will put RB and BB
consecutively (because they are essentially indistinguishable).  Can
we quantify the strength of this effect?  To do this, we define a
bespoke function \code{RB\_BB\_LF()} which returns a \code{hyper3}
log-likelihood function corresponding to repeated observations of our
commuter's reported ranks for the five options:

@@

Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Observe that in almost every case he ranks RB and BB consecutively.
Function \code{RB_BB_LF()} takes argument \code{lambda} that
quantifies the perceived similarity between RB and BB in the same way
as function \code{oldfirm()} above.  It is straightforward to find a
profile support function for $\lambda$,
Figure~\ref{profilesupportlambdaRBBB}, and formally maximize the
likelihood:

\begin{figure}[htbp]
\begin{center}
<<plotproflike>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
lambda <- exp(seq(from=log(0.9), to=log(47), len=17)) 
L <- sapply(lambda, o)
plot(log(lambda), L-max(L), type="b")
abline(h=c(0, -2))
abline(v=0)
@ 
\caption{A\label{profilesupportlambdaRBBB} profile support function
  for the red bus, blue bus parameter $\lambda$.  Vertical line
  indicates null of $\lambda=1$, horizontal lines show maximum support
  and $-2$}
\end{center}
\end{figure}

@@

<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- o(osup$maximum) - o(1))
@

Easily satisfying \citeauthor{edwards1992}'s two-units-of-support
criterion; alternatively, we could observe that this observation lies
in the tail region of its asymptotic $\chi^2_1$ distribution (Wilks's
theorem), with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
  lower.tail=FALSE),3)}.

@@

\begin{figure}[htbp]
  \begin{center}
<<calclikesushi, cache=TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H,give=1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = 12))
L <- sapply(s,f)
plot(log(s),L-max(L),type='b',
     xlab=expression(log(lambda)), ylab='Log-likelihood')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$.  Vertical gray line shows
  $H_0\colon\lambda =1$, horizontal lines show the
  two-units-of-support limits}
\end{center}
\end{figure}

@@

<<defredbusfunc, echo=FALSE>>=
mk <- function(lam){
  H3 <- hyper3()
  for(i in seq_len(nrow(a))){

    d <- strsplit(a$deathorder[i],", ")[[1]]
    H3 <- H3 + cheering3(d, e=ec, help=c(1,lam))
  }
  return(H3)
}
f <- function(l){maxp(mk(l), give=1)$`Log-likelihood`}
@

<<findopt, cache=TRUE, echo=FALSE>>=
optlam <- optimize(f,c(0.09, 0.14), maximum = TRUE)
evaluate <- optlam$maximum
support <- optlam$objective - f(1)
pvalue <- pchisq(support,df=1,lower.tail=FALSE)
@

<<calcklikelam, cache=TRUE, echo=FALSE>>=
lam <- exp(seq(from = -4,to = log(1.1), len = 23))
L <- sapply(lam, f)
@

@@

\begin{figure}[htbp]
  \begin{center}
<<plotlikelam, echo=FALSE>>=
plot(log(lam), L - max(L), type='b')
segments(x0=log(optlam$maximum),y0 = -0.5, y1 = 0.1,col='red')
abline(h=c(0,-2), col = 'gray')
segments(x0 = log(jj1$root), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(jj2$root), y0 = -2.2, y1 = -1.8, col = 'blue')
abline(v = 0, lty = 2)
@ 
\caption{A\label{profsupcounterstrike} profile support function for
  the red bus-blue bus parameter $\log\lambda$ applied to mutually
  supporting equivalence class
  $\left\lbrace\mbox{\skad},\mbox{\stew}\right\rbrace$.  Vertical gray
  line shows $H_0\colon\lambda =1$, for which the support is about
  $\Sexpr{round(support,2)}$ relative to the evaluate
  $\hat{\lambda}\simeq\Sexpr{round(evaluate,3)}$ (small red bar),
  corresponding to a $p$-value of about $\Sexpr{round(pvalue,3)}$.
  Horizontal lines show the two-units-of-support credible interval of
  $(\Sexpr{round(lowerlimit,3)},\Sexpr{round(upperlimit,3)})$, marked
  in blue}
\end{center}
\end{figure}

>>>>

About 22\% of matches in Table~\ref{oldfirm} end in a draw and {\tt
  hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this has a
number of desirable features not present in the reified entity
approach of~\citet{hankin2020}.  Firstly, if we consider two players
of identical strengths and nullify the home ground effect (that is,
$\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
odds of a draw is simply $D$, regardless of team strength.  Secondly,
considering only won (that is, not score-drawn) games, the probability
of a home win is $\frac{\lambda}{1+\lambda}$, that is, the odds of a
home win is simply $\lambda$, independently of the player
strength. Such arguments give direct operational significance to
$\lambda$ and $D$.

\begin{table}
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + \lambda p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
\end{table}

@@

<<makecontourdata, cache=TRUE>>=
n <- 20
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- pmax(-20, LL - max(LL))
@

<<>>=
lambda
@

\begin{figure}[htbp]
  \begin{center}
<<docontourplot, cache=FALSE>>=
contour(lambda, D, matrix(LL, n, n),levels = -2*(1:9), xlab=expression(lambda),ylab=expression(D))
points(maxv[1], maxv[2], pch=16, col='red')
@
\caption{Contours of support\label{contourlambdaD} evaluated over
  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
  a red dot at the evaluate}
  \end{center}
\end{figure}

@@

The classic red bus-blue bus problem was originally stated by
\cite{mcfadden1980} although different metaphors have been employed
for the same phenomenon by \cite{tverskey1972}.  The canonical problem
is as follows.  A traveller has a choice of going to work by different
transport methods.  Here, we can assume that the traveller has strong
but unknown views on different forms of transport which might
incorporate a random component.  If considering whether to use a car
or a bus, the colour of the bus is of no importance to him.  Such
preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, say, $p_{C} = p_{RB}$ and also
indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\Prob(C| \left\lbrace C,RB\right\rbrace)=\frac{1}{3}$.  Various
solutions to this problem have been proposed.  \cite{tverskey1972},
for example, presents an approach involving overlapping ``aspects'' of
the elements of choice set; \cite{mcfadden1980} suggests using a
generalized extreme-value model.

Here I sugest an alternative in the form of a generalization of
Plackett-Luce likelihood functions for ranking options, using the
multiplicative generalization to Bradley-Terry strengths presented
above.  If we ask our respondent to rank his options, it is highly
probable that he will put RB and BB consecutively (because they are
essentially indistinguishable).  Can we quantify the strength of this
effect?  To do this, we define a bespoke function \code{RB\_BB\_LF()}
which returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options:

@@

Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Observe that in almost every case he ranks RB and BB consecutively.
Function \code{RB_BB_LF()} takes argument \code{lambda} that
quantifies the perceived similarity between RB and BB in the same way
as function \code{oldfirm()} above.

<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@

Above, we define function \code{o()} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \code{o()} as a
support function for $\lambda$, and then maximize it:

@@

<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@

We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
  lower.tail=FALSE),3)}.  It is straightforward to find a profile
support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
We see that support of $\log\lambda$ is closely quadratic over a broad
range.

\begin{figure}[htbp]
\begin{center}
<<plotproflike>>=
lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
L <- sapply(lambda, o)
plot(log(lambda), L-max(L), type="b")
abline(h=c(0, -2))
abline(v=0)
@ 
\caption{A\label{profilesupportlambdaRBBB} profile support function
  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
  Vertical line indicates null of $\lambda=1$, horizontal lines show
  maximum support and $-2$}
\end{center}
\end{figure}

@@

<<calclikesushi,echo=FALSE, cache=TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H,give=1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = 23))
L <- sapply(s,f)
@

\begin{figure}[htbp]
  \begin{center}
<<plotlikesushi>>=
plot(log(s),L-max(L),type='b',
     xlab=expression(log(lambda)), ylab='Log-likelihood')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$ in the case of sushi preferences.
  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
  show the two-units-of-support limits}
\end{center}
\end{figure}

@@

<<defredbusfunc, echo=FALSE>>=
mk <- function(lam){
    e <- c(Skadoodle=2, RUSH=1, tarik=1, autimatic=1, Stewie2K=2)
    H3 <- hyper3()
    for(i in seq_len(nrow(a))){
        d <- strsplit(a$deathorder[i],", ")[[1]]
        H3 <- H3 + cheering3(d, e=e, help=c(1, lam))
    }
    return(H3)
}
f <- function(l){maxp(mk(l), give=1)$`Log-likelihood`}
@

<<findopt, cache=TRUE, echo=TRUE>>=
optlam <- optimize(f,c(0.09, 0.14), maximum = TRUE)
evaluate <- optlam$maximum
support <- optlam$objective - f(1)
pvalue <- pchisq(support, df = 1, lower.tail = FALSE)
@

<<calcklikelambda, cache=TRUE, echo=FALSE>>=
lam <- exp(seq(from = -4,to = log(1.1), len = 23))
L <- sapply(lam, f)
@

@@

\begin{figure}[htbp]
  \begin{center}
<<plotlikelam, echo=FALSE>>=
plot(log(lam), L - max(L), type='b')
segments(x0=log(optlam$maximum),y0 = -0.5, y1 = 0.1,col='red')
abline(h=c(0,-2), col = 'gray')
segments(x0 = log(lowerlimit), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(upperlimit), y0 = -2.2, y1 = -1.8, col = 'blue')
abline(v = 0, lty = 2)
@ 
\caption{A\label{profsupcounterstrike} profile support function for
  the red bus-blue bus parameter $\log\lambda$ applied to mutually
  supporting equivalence class
  $\left\lbrace\mbox{\skad},\mbox{\stew}\right\rbrace$.  Vertical gray
  line shows $H_0\colon\lambda =1$, for which the support is about
  $\Sexpr{round(support,2)}$ relative to the evaluate
  $\hat{\lambda}\simeq\Sexpr{round(evaluate,3)}$ (small red bar),
  corresponding to a $p$-value of about $\Sexpr{round(pvalue,3)}$.
  Horizontal lines show the two-units-of-support credible interval of
  $(\Sexpr{round(lowerlimit,3)},\Sexpr{round(upperlimit,3)})$, marked
  in blue}
\end{center}
\end{figure}
